---
title: "Merge Coordinate Updates into NPS Water Systems Database"
author: "Abdullah"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE, comment = "#>")
```

# Overview

Merges coordinate updates from three working copies of the NPS Water Systems
Database into the original workbook, preserving the original's formatting
(merged category header, column widths, styles).

**Merge rule** (applied per coordinate — longitude and latitude are evaluated
independently):

| Original             | Working file          | Action    |
|----------------------|-----------------------|-----------|
| `NA` / `""` / `"NA"` | real number           | ✅ update |
| `NA` / `""` / `"NA"` | `"U"`                 | ✅ update |
| `"U"`                | real number           | ✅ update |
| real number          | different real number | ✅ update |
| anything else        | anything else         | ⏭ skip   |

In words: empty originals accept anything (including `"U"` as a "reviewed
but unknown" marker). A `"U"` upgrades only to a real number. A real number
is never downgraded. When either coord qualifies, the **whole row** is
written from the working file into the original.

**Row matching:** composite key of `park_unit + wsd_system_id + wsd_source_id`.
Guards against `wsd_source_id` typos and cross-park `_01` collisions.

**Conflict rule:** if two working files update the same row, the file listed
*later* in `working_files` wins.

**Outputs (alongside the merged xlsx):**

- `merge_change_log.csv` — every overwritten row, with before/after coords
- `duplicate_source_ids_in_original.csv` — only if duplicate IDs exist
- `rows_needing_coordinates.csv` — only if some rows have `NA` coords across the original *and* all working files

---

# 1. Setup

```{r libs}
library(openxlsx)   # round-trip xlsx while preserving formatting
library(readxl)     # fast text-safe reading
library(dplyr)
```

```{r paths}
# ---- Load per-machine paths from config.R -----------------------------------
# config.R is NOT in git. Each user keeps their own copy with paths to
# wherever their data lives. If you don't have one yet, copy
# config.example.R to config.R and edit the paths inside.
#
# config.R must define: original_file, working_files, output_file
if (!file.exists("config.R")) {
  stop("config.R not found in the working directory. ",
       "Copy config.example.R to config.R and edit the paths inside.")
}
source("config.R")

# ---- Sanity check: did config.R define what we need? ------------------------
required_vars <- c("original_file", "working_files", "output_file")
missing_vars  <- setdiff(required_vars, ls())
if (length(missing_vars) > 0) {
  stop("config.R is missing variable(s): ", paste(missing_vars, collapse = ", "))
}

# ---- Workbook structure (change only if the schema changes) -----------------
sheet_name <- "NPS_Water_Systems_Database"
header_row <- 2   # row 1 = merged category labels ("Park Info", "Source", ...)
                  # row 2 = real column names (park_name, region, ...)
key_cols   <- c("park_unit", "wsd_system_id", "wsd_source_id")

# ---- Derived paths: all diagnostic CSVs land next to the merged xlsx --------
output_dir      <- dirname(output_file)
change_log_path <- file.path(output_dir, "merge_change_log.csv")
dup_path        <- file.path(output_dir, "duplicate_source_ids_in_original.csv")
needs_path      <- file.path(output_dir, "rows_needing_coordinates.csv")

if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
stopifnot(file.access(output_dir, 2) == 0)  # fail fast if output_dir isn't writable
cat("Outputs will be written to:", normalizePath(output_dir, winslash = "/"), "\n")
```

```{r helpers}
# ------------------------------------------------------------------------------
# Composite key: handles typos like AGFO_WS_005 having wsd_source_id
# AGFO_WS_006_01 -- if park_unit + system_id + source_id all match, it's the
# same row, even if an internal reference is wrong.
# ------------------------------------------------------------------------------
make_key <- function(df) {
  paste(
    trimws(as.character(df$park_unit)),
    trimws(as.character(df$wsd_system_id)),
    trimws(as.character(df$wsd_source_id)),
    sep = "||"
  )
}

# ------------------------------------------------------------------------------
# Excel stores dates as serial numbers (days since 1899-12-30). When we read
# with col_types="text", readxl hands us the raw serial (e.g. "46112") instead
# of the displayed string "4/13/2026". If we then write those serials back,
# they appear as 5-digit numbers in the output. Fix: detect serials and
# convert them to M/D/YYYY strings. Leaves already-formatted dates, "U", "NA"
# untouched.
#
# Add column names here if you find another date column that needs this fix.
# ------------------------------------------------------------------------------
date_cols <- c("db_date", "pod_date")

looks_like_excel_serial <- function(x) {
  s <- trimws(as.character(x))
  n <- suppressWarnings(as.numeric(s))
  # modern Excel dates: ~20000 (1954) to ~80000 (2119); must be ~integer so we
  # don't mistake coordinates like -103.473 for dates
  !is.na(n) & n >= 20000 & n <= 80000 & abs(n - round(n)) < 1e-6
}

fix_excel_date_col <- function(x) {
  s <- trimws(as.character(x))
  is_serial <- looks_like_excel_serial(s)
  if (any(is_serial)) {
    # 1899-12-30 origin correctly handles Excel's 1900 leap-year bug
    dates <- as.Date(as.numeric(s[is_serial]), origin = "1899-12-30")
    s[is_serial] <- format(dates, "%m/%d/%Y")
    # Strip leading zeros: "06/04/2025" -> "6/4/2025" (matches Excel default)
    s[is_serial] <- gsub("(^|/)0+([0-9])", "\\1\\2", s[is_serial])
  }
  s
}

fix_date_cols <- function(df, cols = date_cols) {
  for (col in intersect(cols, names(df))) df[[col]] <- fix_excel_date_col(df[[col]])
  df
}

# ------------------------------------------------------------------------------
# Coordinate normalization: round lon/lat to 4 decimal places (~11 m precision
# at the equator). This is well beyond the accuracy of any method we're using
# to locate these wells (Google Earth, building lookups, state databases), so
# no information is lost.
#
# Two things this fixes:
#   1. Floating-point noise from Excel re-saving values (-80.92801 ->
#      -80.928009984783898 etc.) no longer triggers spurious updates.
#   2. Any invisible trailing whitespace (e.g. U+2003 em-space) is silently
#      stripped because as.numeric() ignores it.
#
# "U", "NA", blanks, and other non-numeric values pass through unchanged.
# ------------------------------------------------------------------------------
coord_cols <- c("source_longitude", "source_latitude")
coord_digits <- 4

round_coord <- function(x, digits = coord_digits) {
  s <- trimws(as.character(x))
  n <- suppressWarnings(as.numeric(s))
  ifelse(is.na(n), s, format(round(n, digits), nsmall = 0, trim = TRUE, scientific = FALSE))
}

round_coord_cols <- function(df) {
  for (col in intersect(coord_cols, names(df))) df[[col]] <- round_coord(df[[col]])
  df
}

# ------------------------------------------------------------------------------
# Classifiers for the merge rule.
#
# Blankness is asymmetric between the two sides of the merge:
#   - On the ORIGINAL side we distinguish "never reviewed" (NA/blank) from
#     "reviewed but unknown" (U). The first can accept anything non-blank;
#     the second only upgrades to a real number.
#   - On the WORKING-FILE side we need two predicates: "any non-blank
#     response" (fills an NA) vs "real number" (can upgrade a U or replace
#     a different number).
# ------------------------------------------------------------------------------

# ORIGINAL: truly missing (NA, "", "NA")
is_na_missing <- function(x) is.na(x) | trimws(as.character(x)) %in% c("", "NA")

# ORIGINAL: marked "U" (reviewed, location unknown)
is_U <- function(x) trimws(as.character(x)) %in% c("U", "u")

# WORKING: parses as a real number (excludes "U", "NA", text, blanks)
is_real_coord <- function(x) {
  s <- trimws(as.character(x))
  ok <- !is.na(s) & s != "" & !(s %in% c("NA", "U", "u"))
  suppressWarnings(ok & !is.na(as.numeric(s)))
}

# WORKING: any non-blank value (real number OR "U")
is_non_blank <- function(x) {
  s <- trimws(as.character(x))
  !is.na(s) & s != "" & s != "NA"
}
```

---

# 2. Read the Original

```{r read-original}
# Read everything as text so coordinates (and other numeric strings) aren't
# silently reformatted by readxl's type inference.
orig_df <- read_excel(original_file, sheet = sheet_name, skip = 1, col_types = "text")
orig_df <- fix_date_cols(orig_df)
orig_df <- round_coord_cols(orig_df)

cat("Original rows:", nrow(orig_df), "\n")

# Fail fast if expected columns are missing.
stopifnot(all(c(key_cols, "source_longitude", "source_latitude") %in% names(orig_df)))

# Attach composite key for fast lookups later.
orig_df$.key <- make_key(orig_df)
```

---

# 3. Duplicate Diagnostics

```{r dup-check}
# Two kinds of duplicates to check for:
#   (1) wsd_source_id alone -- informational, tells us if the source_id
#       convention is being reused across parks
#   (2) full composite key -- actionable, because it means our matching will
#       be ambiguous and only the first row will be updated
dup_source_id <- orig_df %>% group_by(wsd_source_id) %>% filter(n() > 1) %>% ungroup()
dup_composite <- orig_df %>% group_by(.key)          %>% filter(n() > 1) %>% ungroup()

cat("Rows sharing wsd_source_id:", nrow(dup_source_id), "\n")
cat("Rows sharing composite key:", nrow(dup_composite), "\n")

if (nrow(dup_source_id) > 0) {
  write.csv(
    dup_source_id %>% select(park_unit, wsd_system_id, wsd_source_id,
                             source_longitude, source_latitude),
    dup_path, row.names = FALSE
  )
  cat("Wrote", dup_path, "\n")
}

if (nrow(dup_composite) > 0) {
  warning("Composite-key duplicates exist. Only the first row of each duplicated ",
          "key will be updated. Clean these in the original before relying on ",
          "the merge for those specific rows.")
}
```

---

# 4. Load the Workbook

```{r load-wb}
wb <- loadWorkbook(original_file)

# ------------------------------------------------------------------------------
# Strip shared-formula definitions from the target sheet.
#
# WHY: openxlsx can't correctly rewrite cells that are part of a shared-formula
# range. When it tries, Excel flags the output as "needs repair" on open
# (Removed Records: Shared formula...). This is a known limitation.
#
# SAFE BECAUSE: this workbook is a static database, not a live-computed
# spreadsheet. Formula cells already have their evaluated values cached by
# Excel; blanking the formula definitions leaves those cached values intact
# in the displayed sheet.
# ------------------------------------------------------------------------------
strip_shared_formulas <- function(wb, sheet) {
  idx <- which(names(wb) == sheet)
  if (length(idx) == 0) return(invisible(wb))
  sd <- wb$worksheets[[idx]]$sheet_data
  if (!is.null(sd$f)) {
    sd$f <- rep("", length(sd$f))
    wb$worksheets[[idx]]$sheet_data <- sd
  }
  invisible(wb)
}
strip_shared_formulas(wb, sheet_name)

# Accumulators for the merge loop.
change_log <- data.frame(
  region = character(), park_unit = character(),
  wsd_system_id = character(), wsd_source_id = character(),
  source_file = character(),
  old_longitude = character(), old_latitude = character(),
  new_longitude = character(), new_latitude = character(),
  stringsAsFactors = FALSE
)

# Per-file coord snapshots, keyed, used by section 6 to flag rows that are
# NA in the original AND in every working file.
work_coord_snapshots <- list()
```

---

# 5. Merge

```{r merge-loop}
for (wf in working_files) {

  if (!file.exists(wf)) { warning("Missing file skipped: ", wf); next }
  cat("\n--- Processing:", wf, "---\n")

  work_df <- read_excel(wf, sheet = sheet_name, skip = 1, col_types = "text")
  work_df <- fix_date_cols(work_df)
  work_df <- round_coord_cols(work_df)

  # Align to the original's schema. If a working file has an extra column,
  # we ignore it; if it's missing one, we only copy the columns present.
  common_cols <- setdiff(intersect(names(orig_df), names(work_df)), ".key")
  work_df <- work_df[, common_cols, drop = FALSE]

  if (!all(key_cols %in% names(work_df))) {
    warning("  Skipping (missing key columns): ", wf); next
  }
  work_df$.key <- make_key(work_df)

  # Save a keyed snapshot of this file's coords for the post-merge diagnostic.
  work_coord_snapshots[[wf]] <- work_df %>%
    select(.key, source_longitude, source_latitude) %>%
    distinct(.key, .keep_all = TRUE)

  # Warn (but don't stop) if the working file has duplicated keys.
  wdup <- work_df %>% group_by(.key) %>% filter(n() > 1) %>% ungroup()
  if (nrow(wdup) > 0) {
    cat("  WARNING:", nrow(wdup), "duplicate composite keys (using first of each)\n")
  }

  # ---------------------------------------------------------------------------
  # Compute which rows qualify for an update.
  #
  # For each coord we apply the three-state rule from the Overview:
  #   orig_na      & new_any   -> update (fills a truly empty cell)
  #   orig_U       & new_real  -> update (upgrades a U to a real number)
  #   orig_real    & new_real  -> update iff the number changed
  # All other combinations are skipped.
  # ---------------------------------------------------------------------------
  orig_lookup <- orig_df %>%
    select(.key, orig_lon = source_longitude, orig_lat = source_latitude)

  updates <- work_df %>%
    select(.key, source_longitude, source_latitude) %>%
    distinct(.key, .keep_all = TRUE) %>%
    left_join(orig_lookup, by = ".key") %>%
    mutate(
      # classify each side
      orig_lon_na = is_na_missing(orig_lon),   orig_lat_na = is_na_missing(orig_lat),
      orig_lon_U  = is_U(orig_lon),            orig_lat_U  = is_U(orig_lat),
      new_lon_real = is_real_coord(source_longitude),
      new_lat_real = is_real_coord(source_latitude),
      new_lon_any  = is_non_blank(source_longitude),
      new_lat_any  = is_non_blank(source_latitude),

      # per-coord update decision
      lon_update = (orig_lon_na & new_lon_any) |
                   (orig_lon_U  & new_lon_real) |
                   (!orig_lon_na & !orig_lon_U & new_lon_real &
                      trimws(source_longitude) != trimws(orig_lon)),
      lat_update = (orig_lat_na & new_lat_any) |
                   (orig_lat_U  & new_lat_real) |
                   (!orig_lat_na & !orig_lat_U & new_lat_real &
                      trimws(source_latitude) != trimws(orig_lat)),

      # row-level: if EITHER coord qualifies, the whole row is rewritten
      differs = lon_update | lat_update
    ) %>%
    filter(differs)

  cat("  Rows to update:", nrow(updates), "\n")
  if (nrow(updates) == 0) next

  # ---------------------------------------------------------------------------
  # Apply updates one at a time. For each qualifying key we:
  #   1. log old vs. new coords to change_log
  #   2. update orig_df in memory so LATER files compare against the merged
  #      state (this is how "last file wins" works without extra bookkeeping)
  #   3. write the entire row's columns back into the workbook object
  # ---------------------------------------------------------------------------
  for (i in seq_len(nrow(updates))) {
    k <- updates$.key[i]

    df_row <- which(orig_df$.key == k)
    if (length(df_row) == 0) { warning("  Key not in original: ", k); next }
    if (length(df_row) > 1) {
      warning("  Composite key matches ", length(df_row),
              " rows; updating only the first: ", k)
      df_row <- df_row[1]
    }

    new_row_df <- work_df[work_df$.key == k, , drop = FALSE][1, , drop = FALSE]

    # (1) log
    change_log <- rbind(change_log, data.frame(
      region        = orig_df$region[df_row],
      park_unit     = orig_df$park_unit[df_row],
      wsd_system_id = orig_df$wsd_system_id[df_row],
      wsd_source_id = orig_df$wsd_source_id[df_row],
      source_file   = wf,
      old_longitude = as.character(orig_df$source_longitude[df_row]),
      old_latitude  = as.character(orig_df$source_latitude[df_row]),
      new_longitude = as.character(new_row_df$source_longitude),
      new_latitude  = as.character(new_row_df$source_latitude),
      stringsAsFactors = FALSE
    ))

    # (2) update in-memory state so downstream files see the latest values
    for (col in common_cols) orig_df[[col]][df_row] <- new_row_df[[col]]

    # (3) write back to the workbook, one column at a time to avoid type
    # coercion quirks. excel_row accounts for the category row at row 1 and
    # header row at row 2, so data row 1 is excel row 3.
    excel_row <- df_row + header_row
    for (col in common_cols) {
      col_idx <- match(col, names(orig_df))
      writeData(wb, sheet = sheet_name, x = new_row_df[[col]],
                startRow = excel_row, startCol = col_idx, colNames = FALSE)
    }
  }
}
```

---

# 6. Rows Needing Attention

Flags rows where coords are still `NA` in the original **and** in every
working file — i.e., nobody has looked at them yet. A `"U"` anywhere means
a reviewer touched the row and is not flagged here.

```{r untouched-rows}
# Re-read the original from disk so "was NA originally" isn't influenced by
# the in-memory updates we applied during the merge loop.
pristine_orig <- read_excel(original_file, sheet = sheet_name, skip = 1, col_types = "text")
pristine_orig <- fix_date_cols(pristine_orig)
pristine_orig <- round_coord_cols(pristine_orig)
pristine_orig$.key <- make_key(pristine_orig)

check_df <- pristine_orig %>%
  select(.key, park_unit, wsd_system_id, wsd_source_id, water_system_name,
         orig_lon = source_longitude, orig_lat = source_latitude) %>%
  mutate(orig_lon_na = is_na_missing(orig_lon),
         orig_lat_na = is_na_missing(orig_lat))

# Attach each working file's coords as lon__<path> / lat__<path> columns
# so we can test "all files NA" with apply() across those columns.
for (wf in names(work_coord_snapshots)) {
  snap <- work_coord_snapshots[[wf]] %>%
    rename(!!paste0("lon__", wf) := source_longitude,
           !!paste0("lat__", wf) := source_latitude)
  check_df <- left_join(check_df, snap, by = ".key")
}

lon_file_cols <- grep("^lon__", names(check_df), value = TRUE)
lat_file_cols <- grep("^lat__", names(check_df), value = TRUE)

# A coord "needs attention" if the original was NA AND every working file is
# also NA for that coord. We flag a ROW if either coord meets this.
check_df$all_lon_na <- check_df$orig_lon_na &
  apply(check_df[, lon_file_cols, drop = FALSE], 1, function(x) all(is_na_missing(x)))
check_df$all_lat_na <- check_df$orig_lat_na &
  apply(check_df[, lat_file_cols, drop = FALSE], 1, function(x) all(is_na_missing(x)))

needs_attention <- check_df %>%
  filter(all_lon_na | all_lat_na) %>%
  select(park_unit, wsd_system_id, wsd_source_id, water_system_name,
         orig_lon, orig_lat, all_of(lon_file_cols), all_of(lat_file_cols))

cat("Rows still needing coordinates:", nrow(needs_attention), "\n")

if (nrow(needs_attention) > 0) {
  warning(nrow(needs_attention), " row(s) need research. See ", needs_path)
  write.csv(needs_attention, needs_path, row.names = FALSE)
}
```

---

# 7. Save

```{r save}
saveWorkbook(wb, output_file, overwrite = TRUE)
write.csv(change_log, change_log_path, row.names = FALSE)

cat("Merged workbook ->", output_file, "\n")
cat("Change log      ->", change_log_path, "  (", nrow(change_log), "rows updated )\n")
```

---

# Notes

- **Per-coordinate evaluation**: if longitude goes `NA → real` but latitude
  stays `"U"`, the row is still updated. The full row is written back, so
  whatever the working file has for latitude (including `"U"`) comes along.
- **Precedence between working files** = order in `working_files`. The last
  file in the vector wins when two files touch the same row.
- **Net-new rows** present only in a working file are ignored — the matching
  step can't pair them with an original row. If you need to append them,
  that's a separate step.
